---
title: "Random Intercept Models - variance components"
output:
  html_document: default
  html_notebook: default
---


Random intercept models are the basic type of random effect you can add to a linear model to turn it into a generalised linear mixed effects model or GLMM. Essentially, such a model has a fixed part that includes coeffcients that affect the mean of the data and also a random part with coeffcinets that affect the variance: $Y_i \sim dnorm(\mu, \sigma^2)$. These models are also a type of variance partioning, whereby the overall total variance $\sigma^2$ is split into partitions attibutatle to different groupings of the data. This is the basic model we will explore first.

## Variance Partitioning

In this model, we have a nested data structure with three levels: observations nested within groups, nested within a larger community which ultimately together form a population (or some such structure of cells/tissues/organs or individuals/counties/countries etc...). We will simulate the data for this example so we can see how it is formed.

```{r sim-var-part}
rm(list=ls())
graphics.off()
set.seed(2)

library(viridis)

n.obs    <- 4
n.groups <- 24
n.comms  <- 6
n.total  <- n.obs * n.groups * n.comms

# standard deviations for each level
s.groups <- 2
s.comms  <- 18.5
s <- 4.2 # residual error

# calculate the random effects for each grouping variable
V.groups <- rnorm(n.groups, 0, s.groups)
V.comms  <- rnorm(n.comms,  0, s.comms)

# allocate each observation to a category for groups and communities
G <- data.frame(groups = sort(rep(1:n.groups, n.obs*n.comms)),
                comms  = sort(rep(1:n.comms, n.obs*n.groups)))




# grand mean
b0 <- 23

# calculate the mean
mu <- b0 + V.groups[G$groups] + V.comms[G$comms]

# simulate the data at the observation level
Y <- mu + rnorm(n.total, 0, s)

mydata <- data.frame(Y = Y, groups = G$groups, comms = G$comms)

# plot the data
palette(viridis(n.comms))
with(mydata, boxplot(Y ~ groups, col = sort(rep(1:n.comms, n.groups/n.comms))))


```

With the data simluated, we can start modelling it using a variance components model, which is essentially a random effects model with a grand mean, and variance terms for community, group and residual. Remember that we specified our data using means and standard deviations as required by `rnorm(n, mean, sd)` so we need to match up these estimates. And it works!

```{r model-var-part}
library(lme4)

# fit a random effects model
m1 <- lmer(Y ~ 1 + (1|groups) + (1|comms), data = mydata)
summary(m1)

# compare with our specified estimates
print(c(groups = s.groups, comms = s.comms, Residual = s))


# use the nesting notation, which basically deals with a labelling situation if you 
# decide to reuse the group labels within each higher level grouping (communities) 
# in this case. For this example the result is equivalent.
m2 <- lmer(Y ~ 1 + (1|comms/groups), data = mydata)
summary(m2)
```

