---
title: "ggplot2_30_12_2016vr1"
author: "Ruth Kelly"
date: "29 November 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup files and libraries}
setwd("C:/R/Rusers_group_testing/ggplot2rk")

library("ggplot2")
library("datasets")
library("RColorBrewer")
```

## Notes and code for TCD R-users group 
### Ggplo2 package by Hadley Wickham

####  The grammar of graphics - 

##### Core elements:
  1. Data
  2. Aesthetics - aes()
  3. Geoms - geom_()
  4. Themes - theme()
  5. Guides - guide()

##### Data

The information you want to communicate.  This information is generally stored in a data frame, matrix or table

##### Aesthetics 

The aesthetic attributes you use to represent your data.  These translate the information stored in columns in your data frame into visual properties of the plot
such as point location, bar height, colour, size, shape etc..

##### Geoms

These are the geometric objects on the plot, e.g. lines, bars and points

##### Themes 

These adjust the overall appearance of the plot, e.g. background colour

##### Guides

These objects help the viewer to interpret the plot, e.g. axis labels and legends.


###  Let's have a look at some code examples of how these work 

First an example using the freely available Iris dataset. 

```{r iris scatter, echo=TRUE}
data(iris)

iris$abundance <- rpois(nrow(iris), 20)

head(iris)


#### first we set up the plot using the function ggplot and tell the function
#### which data to plot onto the x and y axes 

Piris <- ggplot(iris, aes(x = Sepal.Width, y =  Sepal.Length))

### then we can add different geoms (shapes) to our plot

Piris2 <- Piris + geom_point()
Piris2 

Piris3 <- Piris2 + geom_line()
Piris3 

Piris4 <- Piris2 + geom_smooth(method = 'loess')
Piris4

Piris5 <- Piris2 + geom_smooth(method = 'lm')
Piris5

### Each element added to a plot is refered to as a layer in ggplot 'language'

```

### Grouping objects and coding with colour shape and size etc. 

```{r iris scatter with colour groups, echo=TRUE}

### We can code the points by colours by mapping the column $Species from our 
### data frame to the aesthetic "colour"... as follows 

Piris <- ggplot(iris, aes(x = Sepal.Width, y =  Sepal.Length, colour = Species))
Piris2 <- Piris + geom_point()

Piris3 <- Piris2 + geom_smooth(method = "lm")
Piris3

### We can also adjust sizes, colours, shapes etc. for specific inside geoms.. e.g. 
names(iris)
Piris <- ggplot(iris, aes(x = Sepal.Width, y =  Sepal.Length, colour = Species))
Piris2 <- Piris + geom_point(aes(shape = iris$Species, size = iris$abundance/2,
                                 alpha = 1/10))
Piris2 

### Note: alpha adds transparency.. 
```

#### Now let's try adjusting the plot using "themes"..

```{r barplot example - by Siobhan Porter, echo=TRUE}
Cam <- read.csv("Central_america.csv")
summary(Cam)

### First set up your plot by choosing which variables to show on the x and y axes
### and add geom_bar() which draws bars

p1 <- ggplot(Cam, aes(x = country, y = Density)) + 
  geom_bar(stat = "identity", fill = "light blue", colour = "black")

# stat = "identity" - maps the value of the data to the y aesthetic (the #alternative being 'stat = "bin", which maps the number of cases)
# "fill" = fill colour - can use web colours in R
# "colour" = outline colour

p1

## Change the y label using ylab()
p2 <- p1 + ylab("Population density")

p2

## Add title using ggtitle(). (*km^2 creates superscript 2)
p3 <- p2 + ggtitle(expression(paste("Average population per " *km^2)))

p3

## Add values to the bars - vjust alters the height of the numbers (vertical justification))
p4 <- p3 + geom_text(aes(label=round(Density,1)), vjust=1.2, size = 3)
p4


### We use theme() to change the general look of the plot ####
###  make the title go to the centre of the plot
p5 <- p4 + theme(plot.title = element_text(hjust = 0.5))
p5

## To get rid of the grey background grid
p6 <- p5 + theme(panel.grid.major = element_blank(),
                 panel.grid.minor = element_blank()) +
               theme(panel.background = element_blank())

p6

## get rid of the x axis label
p6 <- p5 + theme(axis.title.x=element_blank())

p6

## Bold face and larger text (+ vjust to move the bigger label away from the plot)

p7 <- p6 + theme(axis.title.y=element_text(face="bold", size=15, vjust=1.5))

p7

## Axis text bold face, black and bigger

p8 <- p7 +  theme(axis.text.x=element_text(face="bold", size=8, colour = "black")) +
  theme(axis.text.y=element_text(face="bold", size=15, colour = "black"))

p8

## Change plot title
p9 <- p8 + theme(plot.title = element_text(face="bold", size=25))

p9


################################
#### All together now.. ####
###############################

p <- ggplot(Cam, aes(x = country, y = Density)) + 
          geom_bar(stat = "identity", fill = "light blue", colour = "black") 

p <- p +  ylab("Population density") +
  ggtitle(expression(paste("Average population per " *km^2))) +
  geom_text(aes(label=Density), vjust=1.2, size = 3) + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(panel.background = element_blank()) +
  theme(axis.title.x=element_blank()) + 
  theme(axis.title.y=element_text(face="bold", size=15, vjust=1.5)) +
  theme(axis.text.x=element_text(face="bold", size=8, colour = "black")) +
  theme(axis.text.y=element_text(face="bold", size=8, colour = "black")) +
  theme(plot.title = element_text(lineheight=.8, face="bold", size=15, hjust =0.5)) 

print(p)
 
```

#### Using "guides" to adjust the legend

An example with a coeffient plot (because I like them)


```{r coefficient plot, echo=TRUE}

All_mods <- read.csv("pollinator_mods.csv")

summary(All_mods)


### Plot of model coefficients
zp1 <- ggplot(All_mods, aes(colour = Species, shape = Species))

zp2 <- zp1 + geom_hline(yintercept = 0, colour = gray(1/4), lty = 2) + 
  geom_pointrange(aes(x = modelName, y = Coefficient, 
                      ymin = Confint_L,
                      ymax = Confint_U),show.legend=T,
                  lwd = 0.5, position = position_dodge(width = 2/3)) + 
  ggtitle("Visiting pollinator abundance") + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  coord_flip() + theme_bw() + 
  theme(axis.title.y=element_blank()) + 
  theme(plot.title = element_text(face="bold", size=15, hjust = 0.5)) + 
   theme(legend.text = element_text(face = "italic", size = 8))  + 
  scale_colour_brewer(palette = "Set1") + 
  theme(axis.text.x = element_text(size = 8)) + 
  theme(axis.text.y = element_text(size = 8)) + 
  theme(axis.title.x = element_text(size = 8)) + 
  theme(plot.title = element_text(size = 11)) + 
  guides(colour = guide_legend(title = NULL))  + 
  guides(shape = guide_legend(title = NULL))

print(zp2)

```

### Colour palettes

Because a lot of people are excited about these..

```{r colour palettes}

display.brewer.all()


```

### Have fun plotting :)  There are many more examples available online!

