---
title: "Heat Maps"
author: "Fiona Roche"
date: "4/18/2018"
#output: ioslides_presentation
output: html_document

---

<style type="text/css">

body{ /* Normal  */
      font-size: 14px;
  }

code.r{ /* Code block */
    font-size: 16px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 16px;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```






## Overview

- How to load and run pheatmap in R
- How to edit the display of your heatmap
- How to add annotations
- How to export elements belonging to each cluster

## How to install and load pheatmap in R

```{r install, echo = TRUE}

# Pheatmap is available through the bioconductor suite
source("http://bioconductor.org/biocLite.R")
biocLite("pheatmap")
biocLite("dplyr")
library("pheatmap") 
library("dplyr")
```


## First create a sample dataset

```{r test, echo = TRUE}

# Example dataset based on Pheatmap tutorial
# https://rdrr.io/cran/pheatmap/man/pheatmap.html
test = matrix(rnorm(200,10,0.75), 20, 10)
test[1:10, seq(1, 10, 2)] = test[1:10, seq(1, 10, 2)] + 3
test[11:20, seq(2, 10, 2)] = test[11:20, seq(2, 10, 2)] + 2
test[15:20, seq(2, 10, 2)] = test[15:20, seq(2, 10, 2)] + 4
test <- as.data.frame(test)
colnames(test) = paste("Test", 1:10, sep = "")
rownames(test) = paste("Gene", 1:20, sep = "")
head(test)
```

## Run pheatmap

```{r heatmap, echo = TRUE}
# Pheatmap in default mode uses Euclidean distance and the "complete method"
pheatmap(test)

```

## Pheatmap using scaling

```{r heatmap5, echo = TRUE}
# Scale by row to view your data relative to the mean of each row. 
# The mean is centred on zero.
# The legend shows Z-scores
pheatmap(test, scale = "row")

```

## Pheatmap using correlation

```{r heatmap4, echo = TRUE}

pheatmap(test, clustering_distance_rows= "correlation")

```


## Editing the display of pheatmap

```{r edit1, echo = TRUE}

# If you want to keep your heatmap in a particular order you can turn off clustering by row. 
# The heatmap will display rows in the order of your dataframe
pheatmap(test, cluster_row = FALSE)

```


```{r edit2, echo = TRUE}

# Remove the legend
pheatmap(test, legend = FALSE)
```


```{r edit3, echo = TRUE}

# Remove row labels, useful if you have a very large heatmap
pheatmap(test, show_rownames = FALSE)
```


```{r edit4, echo = TRUE}

# Show text within cells
pheatmap(test, display_numbers = TRUE)

```

## Add annotations to your heatmap

```{r annotation1, echo = TRUE}

# Create annotations for columns. 
# Very useful for generating pretty heatmaps for publications!
annotation_col = data.frame(
  CellType = factor(rep(c("CT1", "CT2"), 5)),
  Time = 1:5
)
rownames(annotation_col) = paste("Test", 1:10, sep = "")

head(annotation_col)

```


```{r annotation2, echo = TRUE}

# Create annotations for rows
annotation_row = data.frame(
  GeneClass = factor(rep(c("Path1", "Path2", "Path3"), c(10, 4, 6)))
)
rownames(annotation_row) = paste("Gene", 1:20, sep = "")

head(annotation_row)

```


```{r annotation3, echo = TRUE}

# Display column annotations
pheatmap(test, annotation_col = annotation_col)

```



```{r annotation5, echo = TRUE}

# Display row and column annotations
pheatmap(test, annotation_col = annotation_col, 
         annotation_row = annotation_row)

```

## Label individual cells

```{r label1, echo = TRUE}

# Highlight particular cells within your heatmap
labels_row = c("", "", "", "", "", "", "", "", "", "", "", "", "", "", "",
               "", "", "Il10", "Il15", "Il1b")
pheatmap(test, annotation_col = annotation_col,  annotation_row = annotation_row, labels_row = labels_row)
```


## Introduce Gaps

```{r display1, echo = TRUE}

# Add gaps to make your annotation grouping stand out
# You can do this manually by turning the clustering by rows off
# Then you list which rows you would like a gap to be introduced
# If using this approach, your df needs to be ordered how you want it plotted in the heatmap
pheatmap(test, annotation_col = annotation_col, annotation_row = annotation_row, 
         cluster_rows = FALSE, gaps_row = c(10, 14))

```


```{r display2, echo = TRUE}

# You can also add gaps within columns. 
# Here it uses the cutree_col parameter to break the dendrogram into 2 clusters
pheatmap(test, annotation_col = annotation_col,annotation_row = annotation_row,
         cluster_rows = FALSE, gaps_row = c(10, 14), cutree_col = 2)

```



```{r display3, echo = TRUE}

# You can also add gaps at cluster breaks within rows
pheatmap(test, annotation_col = annotation_col, annotation_row = annotation_row,
         cluster_rows = TRUE, cutree_col = 2, cutree_rows = 3)

```


## Export which data points are associated with which clusters

```{r export, echo = TRUE}

# If you have a large heatmap and want to identify which genes are found within each cluster you have generated, you can do the following
# Run pheatmap again but this time read it to a variable. Define the number of clusters you want with cutree_rows (.e.g 3)
result<-pheatmap(test, cluster_cols=TRUE, show_rownames=T, cutree_rows=3)
# Use cutree function to extract the row names associated with each cluster. 
# These clusters are ordered the in the same way they appear in the heatmap graphic
cluster_lists <- cutree(result$tree_row, k=3)[result$tree_row[["order"]]]
cluster_listsDF <- as.data.frame(cluster_lists)
# List your clusters
unique(cluster_listsDF$cluster_lists)
# Set rownames of this df to a variable (e.g. $gene) so you can use the dplyr filter function
cluster_listsDF$gene <- row.names(cluster_listsDF)
head(cluster_listsDF)
# Filter for all genes associated with a particular cluster
gene_subset <- filter(cluster_listsDF, cluster_lists == 3)
gene_subset

```


## Further information on Pheatmap

```{r info, echo=TRUE}

# Pheatmap tutorial

# link to R Package Documentation Site
# https://rdrr.io/cran/pheatmap/man/pheatmap.html

```
